---
title: "analysis"
author: "Tyler Vo"
date: "2026-04-13"
output: pdf_document
---

# Import and Libraries
```{r setup, include=FALSE}
library(tidyr)
library(ggplot2)
library(dplyr)
library(glue)
library(stringr)

raw_df <- read.csv("ref/MUP_INP_RY25_P03_V10_DY23_PrvSvc.CSV")
```

# Roll-up categories
Roll up DRG codes to broad diagnostic categories
```{R}
drg_grouping <- raw_df |>
  mutate(DRG_group = case_when(
    # Neurological
    str_detect(DRG_Desc, "CRANI|NEURO|CEREBR|INTRACRANIAL|SEIZURE|STROKE|NERVOUS") ~ "Neuro",
    # Respiratory
    str_detect(DRG_Desc, "RESPIRATORY|PULMONARY|VENTILATION|LUNG|EDEMA") ~ "Respiratory",
    # Cardiac
    str_detect(DRG_Desc, "CARDIAC|HEART|MYOCARDIAL|CORONARY|ARRHYTHMIA|CHF") ~ "Cardiac",
    # Gastrointestinal / Hepatic
    str_detect(DRG_Desc, "GASTRO|INTESTIN|COLON|LIVER|HEPATIC|PANCREA|DIGEST") ~ "GI/Hepatic",
    # Renal / Urinary
    str_detect(DRG_Desc, "RENAL|KIDNEY|URINARY|BLADDER|DIALYSIS") ~ "Renal/Urinary",
    # Infection / Sepsis
    str_detect(DRG_Desc, "SEPSIS|INFECTION|INFECTIOUS|BACTERIA|VIRAL") ~ "Infectious",
    # Oncology
    str_detect(DRG_Desc, "NEOPLASM|CANCER|LYMPHOMA|LEUKEMIA|MALIGN") ~ "Oncology",
    # Orthopedic / MSK
    str_detect(DRG_Desc, "JOINT|HIP|FEMUR|SPINE|ORTHO|BONE|MUSCULOSKELETAL") ~ "Orthopedic",
    # OB / GYN / Neonatal
    str_detect(DRG_Desc, "PREGNANCY|DELIVERY|CESAREAN|NEWBORN|NEONATE|OBSTETRIC") ~ "OB/GYN",
    # Trauma / Injury
    str_detect(DRG_Desc, "TRAUMA|INJURY|FRACTURE|WOUND|BURN") ~ "Trauma",
    str_detect(DRG_Desc, "PROCEDURE|SURGERY") ~ "Procedures",
    TRUE ~ "Other"))

drg_check <- drg_grouping |>
  select(DRG_Cd, DRG_Desc, DRG_group) |>
  distinct() |>
  group_by(DRG_group) |>
  count() |>
  arrange(desc(n))

drg_check
```

Roll up RUCA codes into broad metropolitan categories based on guidance from the US Department of Agriculture
```{r}
metro_grouping <- drg_grouping |>
  mutate(metro_pre = as.integer(str_extract(as.character(Rndrng_Prvdr_RUCA),"^\\d+")))|>
  mutate(metro_area =case_when(
    metro_pre >= 1 & metro_pre <=3 ~ "Metropolitan",
    metro_pre >=4 & metro_pre <= 6 ~ "Micropolitan",
    metro_pre >= 7 & metro_pre <= 9 ~ "Small town",
    metro_pre == 10 ~ "Rural",
    .default = "Not Coded"
  ))

metro_check <- metro_grouping |>
  select(Rndrng_Prvdr_Org_Name, metro_area) |>
  distinct() |>
  group_by(metro_area) |>
  count() |>
  arrange(n)

```

# Exploratory Data Analysis
How many unique values are there for each variable 
```{r}
df_cols <- names(raw_df) 

print(paste0("There are ", nrow(raw_df), " rows in this dataset"))
  
count_unique <- function(df,var) {
  df |>
    select({{var}}) |>
    distinct() |>
    nrow()
}

for (col in df_cols) {
  print(paste0(col, " has ", count_unique(raw_df, col), " unique values"))
}
```
```{r}
df <- metro_grouping |>
  mutate(avg_patient_pay = Avg_Tot_Pymt_Amt - Avg_Mdcr_Pymt_Amt)

all_discharge <- sum(df$Tot_Dschrgs)
```

Find average cost by relevant categories
```{r}
cost_calc <- function(df, cost_var, group_var) {
  df |>
    group_by(.data[[group_var]]) |>
    mutate(wgt_cost = .data[[cost_var]] * Tot_Dschrgs) |>
    summarise(wgt_cost = sum(wgt_cost),
              wgt_disc = sum(Tot_Dschrgs)) |>
    mutate("{cost_var}" := wgt_cost/wgt_disc) |> # calculate the weighted average
    arrange(desc(wgt_disc))
}

cost_vars <- c("avg_patient_pay", "Avg_Mdcr_Pymt_Amt")
group_vars <- c("Rndrng_Prvdr_Org_Name", "Rndrng_Prvdr_State_Abrvtn", "Rndrng_Prvdr_RUCA_Desc", "DRG_Desc")

for (i in cost_vars) {
  for (j in group_vars){
    tmp <- cost_calc(df, i, j)
    assign(glue("{j}:{i}"), tmp)
  }
}
```
Most inpatient hospital stay costs Medicare less than $25,000. The average Medicare cost per discharge follows a right-skewed normal distribution. This suggests that we should log the Medicare cost column before feeding it into the model.
```{r}
ggplot(`Rndrng_Prvdr_Org_Name:Avg_Mdcr_Pymt_Amt`, aes(x = Avg_Mdcr_Pymt_Amt)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of Medicare spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```
Most patients spend less than $10,000 per discharge. Similarly, the distribution follows a right-skewed normal distribution, suggesting that we should log the patient cost before feeding it into the model. 
```{r}
ggplot(`Rndrng_Prvdr_Org_Name:avg_patient_pay`, aes(x = avg_patient_pay)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of patient spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```

```{r}
drg_group_count <- df |>
  group_by(DRG_group) |>
  summarise(discharge = sum(Tot_Dschrgs)) |>
  mutate(pct = discharge/all_discharge) |>
  arrange(desc(discharge)) |>
  mutate(DRG_group = fct_reorder(DRG_group,pct))

p <- ggplot(drg_group_count, aes(x=DRG_group, y = pct, fill = DRG_group)) +
  geom_col()+
  labs(y = "Discharge Percent") +
  scale_fill_viridis_d(option = "C")
```

The highest cost state is "DC", but that's likely due to higher variance since it has less discharges than any other state.
```{r}

state_patient_oop <- `Rndrng_Prvdr_State_Abrvtn:avg_patient_pay`|>
  mutate(Rndrng_Prvdr_State_Abrvtn = fct_reorder(Rndrng_Prvdr_State_Abrvtn, avg_patient_pay))


ggplot(state_patient_oop, aes(x=Rndrng_Prvdr_State_Abrvtn, y = avg_patient_pay)) +
  geom_col()
```

```{r}
# taking the log of our continuous values so they are closer to a normal distribution
df_clean <- df %>%
  mutate(
    log_spend = log(Avg_Mdcr_Pymt_Amt),
    log_oop = log(avg_patient_pay+.000000001), # to avoid -Inf error.
    log_discharges = log(Tot_Dschrgs),
    log_charge = log(Avg_Submtd_Cvrd_Chrg),
    state = Rndrng_Prvdr_State_Abrvtn
  ) |>
  select(log_spend, log_discharges, log_charge, state, DRG_group, metro_area, Rndrng_Prvdr_Org_Name,log_oop) %>%
  drop_na()

# there's a lot of variation between hospitals. Rolling up variables so our model doesn't capture the noise as much
df_clean <- df_clean |>
  group_by(state, DRG_group, metro_area) |>
  summarise(total_medicare_spend = sum(log_spend),
            total_charge = sum(log_charge),
            total_discharge = sum(log_discharges),
            wgt_avg_medicare_spend = total_medicare_spend/total_discharge,
            wgt_avg_patient_pay = sum(log_oop)/total_discharge,
            wgt_avg_charge = total_charge/total_discharge)

model <- lm(
  wgt_avg_medicare_spend ~ DRG_group + metro_area +state +wgt_avg_charge,
  data = df_clean
)

pay_model <- lm(
  wgt_avg_patient_pay ~ DRG_group + metro_area +state +wgt_avg_charge,
  data = df_clean
)

# Standard summary
summary(model)
summary(pay_model)
```

```{r}
df_clean$patient_pay_rpredict <- predict.lm(pay_model, df_clean)
df_clean$medicare_pay_predict <- predict.lm(model, df_clean)
```

```{R}
##MAE & RMSE Medicare
mean(abs(df_clean$wgt_avg_medicare_spend - df_clean$medicare_pay_predict))
sqrt(mean((df_clean$wgt_avg_medicare_spend - df_clean$medicare_pay_predict)^2))

##MAE & RMSE Patient Pay

mean(abs(df_clean$wgt_avg_patient_pay - df_clean$patient_pay_predict))
sqrt(mean((df_clean$wgt_avg_patient_pay - df_clean$patient_pay_predict)^2))
# ---------------------------
# 5. Diagnostic plots
# ---------------------------
par(mfrow = c(2, 2))
plot(model)
plot(pay_model)
```

plotted some predictions vs actuals

```{r}
ggplot(df_clean, aes(x = patient_pay_rpredict, y = wgt_avg_patient_pay)) +
    geom_point(alpha = .5, color = "black") +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    labs(x = "Predicted", y = "Actual")
    
ggplot(df_clean, aes(x = medicare_pay_predict, y = wgt_avg_medicare_spend)) +
    geom_point(alpha = .5, color = "black") +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    labs(x = "Predicted", y = "Actual")    
```

closed ridge solution for weighted average patient payment, manual lambda = 0.01 
```{r}
ridge_x <- model.matrix(~., data = df_clean[,c(1,2,3,10)])[,-1] #### we do - 1 to get rid of intercept column 
ridge_y <- df_clean$wgt_avg_patient_pay

lambda <- 0.01
I <- diag(ncol(ridge_x))

ridge <- solve(t(ridge_x) %*% ridge_x + lambda*I) %*% t(ridge_x) %*% ridge_y

ridge_yhat <- ridge_x %*% ridge ## matrix multiplicat to get our yhats 
ridge_residuals <- ridge_y - ridge_yhat
ridge_fitted <- ridge_yhat
```

closed ridge solution for weighted average medicare spend, manual lambda = 0.01
```{r}
ridge_x_medicare <- model.matrix(~., data = df_clean[,c(1,2,3,10)])[,-1] #### we do - 1 to get rid of intercept column 
ridge_y_medicare <- df_clean$wgt_avg_medicare_spend

lambda <- 0.01
I <- diag(ncol(ridge_x_medicare))

ridge_medicare <- solve(t(ridge_x_medicare) %*% ridge_x_medicare + lambda*I) %*% t(ridge_x_medicare) %*% ridge_y_medicare

ridge_yhat_m <- ridge_x_medicare %*% ridge_medicare ## matrix multiplicat to get our yhats 
ridge_residuals_m <- ridge_y_medicare - ridge_yhat_m
ridge_fitted_m <- ridge_yhat_m
```
