---
title: "analysis"
author: "Tyler Vo"
date: "2026-04-13"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyr)
library(ggplot2)
library(dplyr)
library(glue)
library(forcats)
library(sandwich)
library(lmtest)
library(stringr)

raw_df <- read.csv("ref/MUP_INP_RY25_P03_V10_DY23_PrvSvc.CSV")
```

How many unique values are there for each variable 
```{r}
df_cols <- names(raw_df) 

print(paste0("There are ", nrow(raw_df), " rows in this dataset"))

count_unique <- function(df,var) {
  df |>
    select({{var}}) |>
    distinct() |>
    nrow()
}

for (col in df_cols) {
  print(paste0(col, " has ", count_unique(raw_df, col), " unique values"))
}
```

```{r}
discrete_vars <- c("Rndrng_Prvdr_Org_Name", "Rndrng_Prvdr_City", "Rndrng_Prvdr_St", "Rndrng_Prvdr_State_FIPS", "Rndrng_Prvdr_Zip5", "Rndrng_Prvdr_State_Abrvtn", "Rndrng_Prvdr_RUCA", "Rndrng_Prvdr_RUCA_Desc", "DRG_Cd", "DRG_Desc")
continuous_vars <- c("Tot_Dschrgs", "Avg_Submtd_Cvrd_Chrg", "Avg_Tot_Pymt_Amt", "Avg_Mdcr_Pymt_Amt")
```

List of questions I have:
- Which hospital has the most inpatient stay?
- Which hospital charges the most/least on average?
  - Could be a good opportunity for a box plot
- Is there a state where IP stays are more costly?
- Which DRG are the most costly for patients? For Medicare?
  - Are they somewhat similar?
  - Does Medicare cover certain procedures at a higher rate than other procedures?
- How much more expensive it MCC vs CC?
  
What we can't answer with this data:
-  Rate of IP stays by state. We will require the total Part A Beneficiaries by State
- No point in doing ANOVAs with this data. The total discharge is extremely high, so any difference in cost is bound to cause the p values to be below .05. Also, ANOVAs only tell us if there exist a difference in means between groups


```{r}
df <- raw_df |>
  mutate(avg_patient_pay = Avg_Tot_Pymt_Amt - Avg_Mdcr_Pymt_Amt)

all_discharge <- sum(df$Tot_Dschrgs)

df_summary <- df |>
  group_by(Rndrng_Prvdr_Org_Name) |>
  summarise(discharge_by_hospital = sum(Tot_Dschrgs)) |>
  mutate(pct_of_all_discharged = discharge_by_hospital/all_discharge) |>
  arrange(desc(discharge_by_hospital))

ggplot(df_summary, aes(x = pct_of_all_discharged)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of % of Total Discharges by Hospital",
    x = "Percent of Total Discharges",
    y = "Density"
  )
```
Based on the graph above, it looks like most hospitals account for less than .1% of the Medicare population

Which hospitals charges the most and least on average?
```{r}
avg_cost_calc <- function(df, var) {
  df |>
  group_by(Rndrng_Prvdr_Org_Name) |>
  summarise(avg_cost = mean(.data[[var]], na.rm = TRUE)) |>
  arrange(desc(avg_cost))
}

medicare_spend_by_hosp <- avg_cost_calc(df, "Avg_Mdcr_Pymt_Amt")

ggplot(medicare_spend_by_hosp, aes(x = avg_cost)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of Medicare spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```
The average hospital stay costs Medicare less than $25,000. For the most part, it follows a normal distribution

```{r}
patient_oop_by_hosp <- avg_cost_calc(df, "avg_patient_pay")

ggplot(patient_oop_by_hosp, aes(x = avg_cost)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of patient spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```
Because we are dealing with medical costs, there are a decent amount of outliers that will skew the average cost to be higher than it should be. It's probably best to use box plots when describing costs.

What is the most common DRG?
```{r}
df |>
  group_by(DRG_Desc) |>
  summarise(discharge = sum(Tot_Dschrgs)) |>
  mutate(pct = discharge/all_discharge) |>
  arrange(desc(discharge))
```

Find average cost by relevant categories
```{r}
cost_calc <- function(df, cost_var, group_var) {
  df |>
    group_by(.data[[group_var]]) |>
    mutate(wgt_cost = .data[[cost_var]] * Tot_Dschrgs) |>
    summarise(wgt_cost = sum(wgt_cost),
              wgt_disc = sum(Tot_Dschrgs)) |>
    mutate("{cost_var}" := wgt_cost/wgt_disc) |>
    arrange(desc(wgt_disc))
}

cost_vars <- c("avg_patient_pay", "Avg_Mdcr_Pymt_Amt")
group_vars <- c("Rndrng_Prvdr_Org_Name", "Rndrng_Prvdr_State_Abrvtn", "Rndrng_Prvdr_RUCA_Desc", "DRG_Desc")

for (i in cost_vars) {
  for (j in group_vars){
    tmp <- cost_calc(df, i, j)
    assign(glue("{j}:{i}"), tmp)
    print(tmp)
  }
}

```

The highest cost state is "DC", but that's likely due to higher variance since it has less discharges than any other state.
```{r}

state_patient_oop <- `Rndrng_Prvdr_State_Abrvtn:avg_patient_pay`|>
  mutate(Rndrng_Prvdr_State_Abrvtn = fct_reorder(Rndrng_Prvdr_State_Abrvtn, avg_patient_pay))


ggplot(state_patient_oop, aes(x=Rndrng_Prvdr_State_Abrvtn, y = avg_patient_pay)) +
  geom_col()
```

```{r}
# taking the log of our continuous values so they are closer to a normal distribution
df_clean <- df %>%
  mutate(
    log_spend = log(Avg_Mdcr_Pymt_Amt),
    log_oop = log(avg_patient_pay+.000000001), # to avoid -Inf error.
    log_discharges = log(Tot_Dschrgs),
    log_charge = log(Avg_Submtd_Cvrd_Chrg),
    state = Rndrng_Prvdr_State_Abrvtn,
    metro_area = Rndrng_Prvdr_RUCA_Desc
  ) %>%
  mutate(DRG_group = case_when(
      str_detect(DRG_Desc, "HEART|CARDIAC") ~ "Cardiovascular",
      str_detect(DRG_Desc, "LUNG|RESPIRATORY") ~ "Respiratory",
      str_detect(DRG_Desc, "KIDNEY|RENAL") ~ "Renal",
      str_detect(DRG_Desc, "SEPSIS|INFECTION") ~ "Infectious",
      str_detect(DRG_Desc, "HIP|FEMUR|JOINT") ~ "Orthopedic",
      TRUE ~ "Other"
    )) |>
  select(log_spend, log_discharges, log_charge, state, DRG_group, metro_area, Rndrng_Prvdr_Org_Name,log_oop) %>%
  drop_na()

# there's a lot of variation between hospitals. Rolling up variables so our model doesn't capture the noise as much
df_clean <- df_clean |>
  group_by(state, DRG_group, metro_area) |>
  summarise(total_medicare_spend = sum(log_spend),
            total_charge = sum(log_charge),
            total_discharge = sum(log_discharges),
            wgt_avg_medicare_spend = total_medicare_spend/total_discharge,
            wgt_avg_patient_pay = sum(log_oop)/total_discharge,
            wgt_avg_charge = total_charge/total_discharge)

# ---------------------------
# 3. Fit regression model
# ---------------------------
model <- lm(
  wgt_avg_medicare_spend ~ DRG_group + metro_area +state +wgt_avg_charge,
  data = df_clean
)

pay_model <- lm(
  wgt_avg_patient_pay ~ DRG_group + metro_area +state +wgt_avg_charge,
  data = df_clean
)

# Standard summary
summary(model)
summary(pay_model)

# ---------------------------
# 4. Robust standard errors
# ---------------------------
coeftest(model, vcov = vcovHC(model, type = "HC1"))

# ---------------------------
# 5. Diagnostic plots
# ---------------------------
par(mfrow = c(2, 2))
plot(model)
plot(pay_model)
```
closed ridge solution for weighted average patient payment, manual lambda = 0.01 
```{r}
ridge_x <- model.matrix(~., data = df_clean[,c(1,2,3,10)])[,-1] #### we do - 1 to get rid of intercept column 
ridge_y <- df_clean$wgt_avg_patient_pay

lambda <- 0.01
I <- diag(ncol(ridge_x))

ridge <- solve(t(ridge_x) %*% ridge_x + lambda*I) %*% t(ridge_x) %*% ridge_y

ridge_yhat <- ridge_x %*% ridge ## matrix multiplicat to get our yhats 
ridge_residuals <- ridge_y - ridge_yhat
ridge_fitted <- ridge_yhat
```

closed ridge solution for weighted average medicare spend, manual lambda = 0.01
```{r}
ridge_x_medicare <- model.matrix(~., data = df_clean[,c(1,2,3,10)])[,-1] #### we do - 1 to get rid of intercept column 
ridge_y_medicare <- df_clean$wgt_avg_medicare_spend

lambda <- 0.01
I <- diag(ncol(ridge_x_medicare))

ridge_medicare <- solve(t(ridge_x_medicare) %*% ridge_x_medicare + lambda*I) %*% t(ridge_x_medicare) %*% ridge_y_medicare

ridge_yhat_m <- ridge_x_medicare %*% ridge_medicare ## matrix multiplicat to get our yhats 
ridge_residuals_m <- ridge_y_medicare - ridge_yhat_m
ridge_fitted_m <- ridge_yhat_m
```

Lasso
```{r}
library(glmnet)


df_clean <- df %>%
  mutate(
    log_spend = log(Avg_Mdcr_Pymt_Amt),
    log_discharges = log(Tot_Dschrgs),
    log_charge = log(Avg_Submtd_Cvrd_Chrg),
    state = as.factor(Rndrng_Prvdr_State_Abrvtn),
    DRG = as.factor(DRG_Cd)
  ) %>%
  select(log_spend, log_discharges, log_charge, state, DRG) %>%
  drop_na()

# ---------------------------
# 2. Create design matrix
# ---------------------------
# glmnet needs a matrix (handles factors automatically)
x <- model.matrix(log_spend ~ ., data = df_clean)[, -1]  # remove intercept
y <- df_clean$log_spend

# ---------------------------
# 3. Train/test split
# ---------------------------
set.seed(123)
train_idx <- sample(nrow(x), 0.8 * nrow(x))

x_train <- x[train_idx, ]
y_train <- y[train_idx]

x_test <- x[-train_idx, ]
y_test <- y[-train_idx]

# ---------------------------
# 4. Fit LASSO (cross-validated)
# ---------------------------
cv_model <- cv.glmnet(
  x_train,
  y_train,
  alpha = 1  # LASSO
)

# Best lambda
best_lambda <- cv_model$lambda.min

# ---------------------------
# 5. Final model
# ---------------------------
lasso_model <- glmnet(
  x_train,
  y_train,
  alpha = 1,
  lambda = best_lambda
)

# ---------------------------
# 6. Coefficients (selected features)
# ---------------------------
coef(lasso_model)

# ---------------------------
# 7. Evaluate performance
# ---------------------------
pred <- predict(lasso_model, s = best_lambda, newx = x_test)

# Mean Squared Error
mean((pred - y_test)^2)
```
