---
title: "analysis"
author: "Tyler Vo"
date: "2026-04-13"
output: pdf_document
---

# Import and Libraries
```{r setup, include=FALSE}
library(tidyr)
library(ggplot2)
library(dplyr)
library(glue)
library(stringr)

raw_df <- read.csv("ref/MUP_INP_RY25_P03_V10_DY23_PrvSvc.CSV")
```

# Roll-up categories
Roll up DRG codes to broad diagnostic categories
```{R}
drg_grouping <- raw_df |>
  mutate(DRG_group = case_when(
    str_detect(DRG_Desc, "CRANI|NEURO|CEREBR|INTRACRANIAL|SEIZURE|STROKE|NERVOUS") ~ "Neuro",
    str_detect(DRG_Desc, "RESPIRATORY|PULMONARY|VENTILATION|LUNG|EDEMA") ~ "Respiratory",
    str_detect(DRG_Desc, "CARDIAC|HEART|MYOCARDIAL|CORONARY|ARRHYTHMIA|CHF") ~ "Cardiac",
    str_detect(DRG_Desc, "GASTRO|INTESTIN|COLON|LIVER|HEPATIC|PANCREA|DIGEST") ~ "GI/Hepatic",
    str_detect(DRG_Desc, "RENAL|KIDNEY|URINARY|BLADDER|DIALYSIS") ~ "Renal/Urinary",
    str_detect(DRG_Desc, "SEPSIS|INFECTION|INFECTIOUS|BACTERIA|VIRAL") ~ "Infectious",
    str_detect(DRG_Desc, "NEOPLASM|CANCER|LYMPHOMA|LEUKEMIA|MALIGN") ~ "Oncology",
    str_detect(DRG_Desc, "JOINT|HIP|FEMUR|SPINE|ORTHO|BONE|MUSCULOSKELETAL") ~ "Orthopedic",
    str_detect(DRG_Desc, "PREGNANCY|DELIVERY|CESAREAN|NEWBORN|NEONATE|OBSTETRIC") ~ "OB/GYN",
    str_detect(DRG_Desc, "TRAUMA|INJURY|FRACTURE|WOUND|BURN") ~ "Trauma",
    str_detect(DRG_Desc, "PROCEDURE|SURGERY") ~ "Procedures",
    TRUE ~ "Other"))

drg_check <- drg_grouping |>
  select(DRG_Cd, DRG_Desc, DRG_group) |>
  distinct() |>
  group_by(DRG_group) |>
  count() |>
  arrange(desc(n))

drg_check
```

Roll up RUCA codes into broad metropolitan categories based on guidance from the US Department of Agriculture
```{r}
metro_grouping <- drg_grouping |>
  mutate(metro_pre = as.integer(str_extract(as.character(Rndrng_Prvdr_RUCA),"^\\d+")))|>
  mutate(metro_area =case_when(
    metro_pre >= 1 & metro_pre <=3 ~ "Metropolitan",
    metro_pre >=4 & metro_pre <= 6 ~ "Micropolitan",
    metro_pre >= 7 & metro_pre <= 9 ~ "Small town",
    metro_pre == 10 ~ "Rural",
    .default = "Not Coded"
  ))

metro_check <- metro_grouping |>
  select(Rndrng_Prvdr_Org_Name, metro_area) |>
  distinct() |>
  group_by(metro_area) |>
  count() |>
  arrange(n)

```

# Exploratory Data Analysis
How many unique values are there for each variable 
```{r}
df_cols <- names(raw_df) 

print(paste0("There are ", nrow(raw_df), " rows in this dataset"))
  
count_unique <- function(df,var) {
  df |>
    select({{var}}) |>
    distinct() |>
    nrow()
}

for (col in df_cols) {
  print(paste0(col, " has ", count_unique(raw_df, col), " unique values"))
}
```
```{r}
df <- metro_grouping |>
  mutate(avg_patient_pay = Avg_Tot_Pymt_Amt - Avg_Mdcr_Pymt_Amt)

all_discharge <- sum(df$Tot_Dschrgs)
```

Find average cost by relevant categories
```{r}
cost_calc <- function(df, cost_var, group_var) {
  df |>
    group_by(.data[[group_var]]) |>
    mutate(wgt_cost = .data[[cost_var]] * Tot_Dschrgs) |>
    summarise(wgt_cost = sum(wgt_cost),
              wgt_disc = sum(Tot_Dschrgs)) |>
    mutate("{cost_var}" := wgt_cost/wgt_disc) |> # calculate the weighted average
    arrange(desc(wgt_disc))
}

cost_vars <- c("avg_patient_pay", "Avg_Mdcr_Pymt_Amt")
group_vars <- c("Rndrng_Prvdr_Org_Name", "Rndrng_Prvdr_State_Abrvtn", "metro_area", "DRG_group")

for (i in cost_vars) {
  for (j in group_vars){
    tmp <- cost_calc(df, i, j)
    assign(glue("{j}:{i}"), tmp)
    print(tmp)
  }
}
```
```{r}
pre_avg <- df |>
  mutate(medicare_wgt = Tot_Dschrgs*Avg_Mdcr_Pymt_Amt,
         oop_wgt = Tot_Dschrgs*avg_patient_pay)
medicare_avg <- sum(pre_avg$medicare_wgt, na.rm = TRUE)/sum(df$Tot_Dschrgs, na.rm=TRUE)
oop_avg <- sum(pre_avg$oop_wgt, na.rm = TRUE)/sum(df$Tot_Dschrgs, na.rm=TRUE)
print(glue("The average discharge costs Medicare ${round(medicare_avg,2)}"))
print(glue("The average patient OOP per discharge costs ${round(oop_avg,2)}"))
print(glue("This means that Medicare pays {round(medicare_avg/oop_avg,2)} times the patient OOP cost on average."))
```
Most inpatient hospital stay costs Medicare less than $25,000. The average Medicare cost per discharge follows a right-skewed normal distribution. This suggests that we should log the Medicare cost column before feeding it into the model.
```{r}
ggplot(`Rndrng_Prvdr_Org_Name:Avg_Mdcr_Pymt_Amt`, aes(x = Avg_Mdcr_Pymt_Amt)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of Medicare spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```
Most patients spend less than $5,000 per discharge. Similarly, the distribution follows a right-skewed normal distribution, suggesting that we should log the patient cost before feeding it into the model. 
```{r}
ggplot(`Rndrng_Prvdr_Org_Name:avg_patient_pay`, aes(x = avg_patient_pay)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of patient spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```

```{r}
drg_group_count <- df |>
  group_by(DRG_group) |>
  summarise(discharge = sum(Tot_Dschrgs)) |>
  mutate(pct = discharge/all_discharge) |>
  arrange(desc(discharge)) |>
  mutate(DRG_group = fct_reorder(DRG_group,pct)) ## this is where it failed. 

p <- ggplot(drg_group_count, aes(x=DRG_group, y = pct, fill = DRG_group)) +
  geom_col()+
  labs(y = "Percent", title = "Percent of Discharge by DRG group") +
  scale_fill_viridis_d(option = "B")
p
```

Oncology and Procedure DRGs are the two most expensive groups on average to patients and Medicare. However, despite patients paying more for oncology care, Medicare reimburses hospitals more for procedures.
```{r}
drg_patient_pay <- `DRG_group:avg_patient_pay` |>
  mutate(DRG_group = fct_reorder(DRG_group,avg_patient_pay, .desc=TRUE))

p <- ggplot(drg_patient_pay, aes(x=DRG_group, y = avg_patient_pay, fill = DRG_group)) +
  geom_col()+
  labs(y = "Average Out-of-Pocket Cost", title = "Average Patient Out-of-Pocket Cost by DRG Group") +
  scale_fill_viridis_d(option = "B")
p
```

```{r}
drg_medicare_cost <- `DRG_group:Avg_Mdcr_Pymt_Amt` |>
  mutate(DRG_group = fct_reorder(DRG_group,Avg_Mdcr_Pymt_Amt, .desc=TRUE))

p <- ggplot(drg_medicare_cost, aes(x=DRG_group, y = Avg_Mdcr_Pymt_Amt, fill = DRG_group)) +
  geom_col()+
  labs(y = "Average Medicare Cost", title = "Average Medicare Cost by DRG Group") +
  scale_fill_viridis_d(option = "B")
p

```
Hospital reimbursement rates are similar across all DRG categories.
```{r}
drg_reimbursement_rate <- drg_medicare_cost |>     ### This failed here. 
  left_join(drg_patient_pay, by = "DRG_group") |>
  mutate(reimbursement_rate = Avg_Mdcr_Pymt_Amt/(Avg_Mdcr_Pymt_Amt+avg_patient_pay)) |>
  mutate(DRG_group = fct_reorder(DRG_group,reimbursement_rate))

p <- ggplot(drg_reimbursement_rate, aes(x=DRG_group, y = reimbursement_rate, fill = DRG_group)) +
  geom_col()+
  labs(y = "Average Medicare Reimbursement Rate", title = "Average Medicare Reimbursement Rate by DRG Group") +
  scale_fill_viridis_d(option = "B")
p
```

```{r}
metro_medicare_cost <- `metro_area:Avg_Mdcr_Pymt_Amt` |>
  mutate(metro_area = fct_reorder(metro_area,Avg_Mdcr_Pymt_Amt, .desc=TRUE))

p <- ggplot(metro_medicare_cost, aes(x=metro_area, y = Avg_Mdcr_Pymt_Amt, fill = metro_area)) +
  geom_col()+
  labs(x = "Metro Area",y = "Average Medicare Cost", title = "Average Medicare Cost by Metro Area") +
  scale_fill_viridis_d(option = "B")
p
```

Medicare reimburses hospitals in metropolitan area at a slightly lower rate compared to all other metropolitan areas.
```{r}
metro_oop_cost <- `metro_area:avg_patient_pay` |>
  mutate(metro_area = fct_reorder(metro_area,avg_patient_pay, .desc=TRUE))

p <- ggplot(metro_oop_cost, aes(x=metro_area, y = avg_patient_pay, fill = metro_area)) +
  geom_col()+
  labs(x = "Metro Area",y = "Average Patient Cost", title = "Average Patient Cost by Metro Area") +
  scale_fill_viridis_d(option = "B")
p
```
```{r}
metro_reimbursement <- left_join(`metro_area:avg_patient_pay`,`metro_area:Avg_Mdcr_Pymt_Amt`, by = "metro_area") |>
  mutate(reimbursement_reate = Avg_Mdcr_Pymt_Amt/(Avg_Mdcr_Pymt_Amt+avg_patient_pay)) |>
  mutate(metro_area = fct_reorder(metro_area,reimbursement_reate, .desc=TRUE))

p <- ggplot(metro_reimbursement, aes(x=metro_area, y = reimbursement_reate, fill = metro_area)) +
  geom_col()+
  labs(x = "Metro Area", y = "Medicare Reimbursement", title = "Average Medicare Reimbursement Rate by Metro Area") +
  scale_fill_viridis_d(option = "B")
p
```

The highest cost state is "DC", but that's likely due to higher variance since it has less discharges than any other state.
```{r}

state_patient_oop <- `Rndrng_Prvdr_State_Abrvtn:avg_patient_pay`|>
  mutate(state_full = tolower(state.name[match(Rndrng_Prvdr_State_Abrvtn, state.abb)])) |>
  mutate(state_full = fct_reorder(state_full, avg_patient_pay))

states_map <- map_data("state")
map_df <- left_join(states_map, state_patient_oop, by = c("region" = "state_full"))

```

```{r}
state_medicare_cost <- `Rndrng_Prvdr_State_Abrvtn:Avg_Mdcr_Pymt_Amt`|>
  mutate(state_full = tolower(state.name[match(Rndrng_Prvdr_State_Abrvtn, state.abb)])) |>
  mutate(state_full = fct_reorder(state_full, Avg_Mdcr_Pymt_Amt))

ggplot(map_df, aes(long, lat, group = group, fill = Avg_Mdcr_Pymt_Amt)) +
  geom_polygon(color = "white") +
  scale_fill_distiller(
  palette = "Blues",
  direction = 1,
  na.value = "grey90"
) +
  theme_void() +
  labs(fill = "Average Medicare Cost", title = "Average Medicare Cost by State")
```

# Modeling
```{r}
# taking the log of our continuous values so they are closer to a normal distribution
df_clean <- df %>%
  mutate(
    log_spend = log(Avg_Mdcr_Pymt_Amt),
    log_oop = log(avg_patient_pay+.000000001), # to avoid -Inf error.
    log_discharges = log(Tot_Dschrgs),
    log_charge = log(Avg_Submtd_Cvrd_Chrg),
    state = Rndrng_Prvdr_State_Abrvtn
  ) |>
  select(log_spend, log_discharges, log_charge, state, DRG_group, metro_area, Rndrng_Prvdr_Org_Name,log_oop) %>%
  drop_na()

# there's a lot of variation between hospitals. Rolling up variables so our model doesn't capture the noise as much
df_clean <- df_clean |>
  group_by(state, DRG_group, metro_area) |>
  summarise(total_medicare_spend = sum(log_spend),
            total_charge = sum(log_charge),
            total_discharge = sum(log_discharges),
            wgt_avg_medicare_spend = total_medicare_spend/total_discharge,
            wgt_avg_patient_pay = sum(log_oop)/total_discharge,
            wgt_avg_charge = total_charge/total_discharge)

model <- lm(
  wgt_avg_medicare_spend ~ DRG_group + metro_area +state +wgt_avg_charge,
  data = df_clean
)

pay_model <- lm(
  wgt_avg_patient_pay ~ DRG_group + metro_area +state +wgt_avg_charge,
  data = df_clean
)

# Standard summary
summary(model)
summary(pay_model)
```

```{r}
df_clean$patient_pay_rpredict <- predict.lm(pay_model, df_clean)
df_clean$medicare_pay_predict <- predict.lm(model, df_clean)
```

```{R}
##MAE & RMSE Medicare
mean(abs(df_clean$wgt_avg_medicare_spend - df_clean$medicare_pay_predict))
sqrt(mean((df_clean$wgt_avg_medicare_spend - df_clean$medicare_pay_predict)^2))

##MAE & RMSE Patient Pay

mean(abs(df_clean$wgt_avg_patient_pay - df_clean$patient_pay_rpredict))
sqrt(mean((df_clean$wgt_avg_patient_pay - df_clean$patient_pay_rpredict)^2))
# ---------------------------
# 5. Diagnostic plots
# ---------------------------
par(mfrow = c(2, 2))
plot(model)
plot(pay_model)
```

plotted some predictions vs actuals

```{r}
ggplot(df_clean, aes(x = patient_pay_rpredict, y = wgt_avg_patient_pay)) +
    geom_point(alpha = .5, color = "black") +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    labs(x = "Predicted", y = "Actual")
    
ggplot(df_clean, aes(x = medicare_pay_predict, y = wgt_avg_medicare_spend)) +
    geom_point(alpha = .5, color = "black") +
    geom_abline(slope = 1, intercept = 0, color = "red") +
    labs(x = "Predicted", y = "Actual")    
```
Ridge setup
```{r}

df_clean$state <- as.factor(df_clean$state)
df_clean$DRG_group <- as.factor(df_clean$DRG_group)
df_clean$metro_area <- as.factor(df_clean$metro_area)


set.seed(1)

training_index <- sample(1:nrow(df_clean),
                         size = 0.8 * nrow(df_clean))

train <- df_clean[training_index,]
test <- df_clean[-training_index,]

ridge_x_train <- model.matrix(~., data = train[,c(1,2,3,9)])[,-1] #### we do - 1 to get rid of intercept column 
ridge_x_test <- model.matrix(~., data = test[,c(1,2,3,9)])[,-1]

ridge_y_train_patient <- train$wgt_avg_patient_pay
ridge_y_test_patient <- test$wgt_avg_patient_pay

ridge_y_train_medicare <- train$wgt_avg_medicare_spend
ridge_y_test_medicare <- test$wgt_avg_medicare_spend



```


Ridge for patient pay
```{r}

ridge_patient <- cv.glmnet(ridge_x_train, ridge_y_train_patient, alpha = 0)
print(ridge_patient$lambda.min)
ridge_patient_predict <-  predict(ridge_patient, s = "lambda.min", newx = ridge_x_test)
test$ridge_patient_predictions <- ridge_patient_predict


```

Ridge for medicare pay. 

```{r}

ridge_medicare <- cv.glmnet(ridge_x_train, ridge_y_train_medicare, alpha = 0)
print(ridge_medicare$lambda.min)
ridge_medicare_predict <- predict(ridge_medicare, s = "lambda.min", newx = ridge_x_test)
test$ridge_medicare_predictions <- ridge_medicare_predict

```

Model analysis:
```{r}
pay_ridge_vector <- as.vector(test$ridge_patient_predictions)
medicare_ridge_vector <- as.vector(test$ridge_medicare_predictions)

## MAE and RMSE Patient

print(mean(abs(test$wgt_avg_patient_pay - test$ridge_patient_predictions)))
print(sqrt(mean((test$wgt_avg_patient_pay - test$ridge_patient_predictions)^2)))

rss_patient <- sum((test$wgt_avg_patient_pay - pay_ridge_vector)^2)
tss_patient <- sum((test$wgt_avg_patient_pay - mean(test$wgt_avg_patient_pay))^2)

r2_patient <- 1 - rss_patient/tss_patient
print(r2_patient)

## MAE and RMSE Medicare

print(mean(abs(test$wgt_avg_medicare_spend - test$ridge_medicare_predictions)))
print(sqrt(mean((test$wgt_avg_medicare_spend - test$ridge_medicare_predictions)^2)))

rss_patient_m <- sum((test$wgt_avg_medicare_spend - medicare_ridge_vector)^2)
tss_patient_m <- sum((test$wgt_avg_medicare_spend - mean(test$wgt_avg_medicare_spend))^2)

r2_medicare <- 1 - rss_patient_m/tss_patient_m
print(r2_medicare)

##R2

```

Comparing the two models:

```{r}

print(sum(test$patient_pay_rpredict - test$ridge_patient_predictions))

ggplot(data = test, aes(x = ridge_patient_predictions, y = patient_pay_rpredict)) + 
    geom_point(alpha = .5) +
    labs ( x = "Ridge Predictions", y = "LM Predictions")
  
## It looks like the lm is predicting higher values compared to the ridge for patient pay, but we have 291 rows so it's not a large difference. 

