---
title: "527 Final Project"
output: pdf_document
---

```{r}
library(dplyr)
library(ggplot2)
df <- read.csv("healthcare_dataset.csv")
```

How many unique values are there for each variable 
```{r}
df_cols <- names(df) 

print(paste0("There are ", nrow(df), " rows in this dataset"))

count_unique <- function(df,var) {
  df |>
    select({{var}}) |>
    distinct() |>
    nrow()
}

for (col in df_cols) {
  print(paste0(col, " has ", count_unique(df, col), " unique values"))
}
```

Preliminary checks:
```{r}
rows_with_na <- !complete.cases(df)
sum(rows_with_na)

rm(rows_with_na)
```
  - Checked type of each variable / column.
```{r}
str(df)
```

Convert date columns into date type, fix names to be First Last, round billing to 2 decimal places 
```{r}
library(lubridate)
library(stringr)
df$Date.of.Admission <- ymd(df$Date.of.Admission)
df$Discharge.Date <- ymd(df$Discharge.Date)
df$Name <- str_to_title(df$Name)
df$Billing.Amount <- round(df$Billing.Amount, digits = 2)

```

Create buckets for age groups
1-10 and 90-100 were intentionally left out as they weren't in the dataset. Please double check that for me though :D 
I also created an age_group vector, not sure if it's needed but could be used to plot, etc 
```{r}
df$Age_Groups <- cut(df$Age, breaks = seq(10,90, by=10), labels = c("11-20", "21-30", "31-40", "41-50", "51-60", "61-70", "71-80", "81-90")) 
age_group_v <- unique(df$Age_Groups)
```

Some summary stats from questions we were looking at in our Descriptive Statistics section. 
```{r}
print((tapply(df$Billing.Amount, 
    list(df$Age_Groups, df$Gender),
    mean)))


print(tapply(df$Billing.Amount, 
    list(df$Age_Groups, df$Gender),
    sd))

print(tapply(df$Medication,
        list(df$Age_Groups, df$Gender),
        function(x) length(unique(x))))

print(tapply(df$Billing.Amount, df$Medical.Condition, mean))

date_diff <- as.numeric(df$Discharge.Date - df$Date.of.Admission)
print(mean(date_diff))
print(sd(date_diff))

print(tapply(df$Test.Results, df$Medical.Condition, length))
```
