---
title: "analysis"
author: "Tyler Vo"
date: "2026-04-13"
output: pdf_document
---

```{r setup, include=FALSE}
library(tidyr)
library(ggplot2)
library(dplyr)
library(glue)

raw_df <- read.csv("ref/MUP_INP_RY25_P03_V10_DY23_PrvSvc.CSV")
```

How many unique values are there for each variable 
```{r}
df_cols <- names(raw_df) 

print(paste0("There are ", nrow(raw_df), " rows in this dataset"))

count_unique <- function(df,var) {
  df |>
    select({{var}}) |>
    distinct() |>
    nrow()
}

for (col in df_cols) {
  print(paste0(col, " has ", count_unique(raw_df, col), " unique values"))
}
```

```{r}
discrete_vars <- c("Rndrng_Prvdr_Org_Name", "Rndrng_Prvdr_City", "Rndrng_Prvdr_St", "Rndrng_Prvdr_State_FIPS", "Rndrng_Prvdr_Zip5", "Rndrng_Prvdr_State_Abrvtn", "Rndrng_Prvdr_RUCA", "Rndrng_Prvdr_RUCA_Desc", "DRG_Cd", "DRG_Desc")
continuous_vars <- c("Tot_Dschrgs", "Avg_Submtd_Cvrd_Chrg", "Avg_Tot_Pymt_Amt", "Avg_Mdcr_Pymt_Amt")
```

List of questions I have:
- Which hospital has the most inpatient stay?
- Which hospital charges the most/least on average?
  - Could be a good opportunity for a box plot
- Is there a state where IP stays are more costly?
- Which DRG are the most costly for patients? For Medicare?
  - Are they somewhat similar?
  - Does Medicare cover certain procedures at a higher rate than other procedures?
- How much more expensive it MCC vs CC?
  
What we can't answer with this data:
-  Rate of IP stays by state. We will require the total Part A Beneficiaries by State


```{r}
df <- raw_df |>
  mutate(avg_patient_pay = Avg_Tot_Pymt_Amt - Avg_Mdcr_Pymt_Amt)

all_discharge <- sum(df$Tot_Dschrgs)

df_summary <- df |>
  group_by(Rndrng_Prvdr_Org_Name) |>
  summarise(discharge_by_hospital = sum(Tot_Dschrgs)) |>
  mutate(pct_of_all_discharged = discharge_by_hospital/all_discharge) |>
  arrange(desc(discharge_by_hospital))

ggplot(df_summary, aes(x = pct_of_all_discharged)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of % of Total Discharges by Hospital",
    x = "Percent of Total Discharges",
    y = "Density"
  )
```
Based on the graph above, it looks like most hospitals have less than .1% of the Medicare Part A population

Which hospitals charges the most and least on average?
```{r}
avg_cost_calc <- function(df, var) {
  df |>
  group_by(Rndrng_Prvdr_Org_Name) |>
  summarise(avg_cost = mean(.data[[var]], na.rm = TRUE)) |>
  arrange(desc(avg_cost))
}

medicare_spend_by_hosp <- avg_cost_calc(df, "Avg_Mdcr_Pymt_Amt")

ggplot(medicare_spend_by_hosp, aes(x = avg_cost)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of Medicare spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```
The average hospital stay costs Medicare less than $25,000. For the most part, it follows a normal distribution

```{r}
patient_oop_by_hosp <- avg_cost_calc(df, "avg_patient_pay")

ggplot(patient_oop_by_hosp, aes(x = avg_cost)) +
  geom_density(fill = "blue", alpha = 0.4) +
  labs(
    title = "Distribution of patient spend by hospitals",
    x = "Total Spend",
    y = "Density"
  )
```
Because we are dealing with medical costs, there are a decent amount of outliers that will skew the average cost to be higher than it should be. It's probably best to use box plots when describing costs.

What is the most common DRG?
```{r}
df |>
  group_by(DRG_Desc) |>
  summarise(discharge = sum(Tot_Dschrgs)) |>
  mutate(pct = discharge/all_discharge) |>
  arrange(desc(discharge))
```
Which DRG is the most expensive for patients?
```{r}
drg_cost_calc <- function(df, var) {
  df |>
    group_by(DRG_Desc) |>
    mutate(wgt_cost = .data[[var]] * Tot_Dschrgs) |>
    summarise(wgt_cost = sum(wgt_cost),
              wgt_disc = sum(Tot_Dschrgs)) |>
    mutate(wgt_avg_cost = wgt_cost/wgt_disc) |>
    arrange(desc(wgt_disc))
}

drg_cost_calc(df, "avg_patient_pay")
```

